---
title: "Accessing and cleaning federal fisheries data"
author: "Carly Lovas"
format:
  html:
    grid: 
      body-width: 2000px
      sidebar-width: 200px
      margin-width: 200px
    toc: true
    self-contained: true
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{data_cleaning_functions}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| message: false
#| warning: false
#| echo: false

library(tidyverse)
library(species.shifts)

box_path <- "/Users/clovas/Library/CloudStorage/Box-Box/MAFMC-25 Data/"
```

## About
This package contains various functions for accessing (pulling) and cleaning federal fisheries dependent and independent data. Many of these data sets are confidential and require a formal data request submitting to the appropriate governing body. Information regarding metadata and access for most of these data sets can be accessed on [inPort](https://www.fisheries.noaa.gov/inport/), see below:

> InPort is the authoritative metadata repository and data inventory platform for NOAA Fisheries and the National Ocean Service. The system supports documentation of datasets and provides tools to facilitate data discovery, public access, and responsible stewardship of scientific data within these line offices.

As such, using `species.shifts` will require access to this data in order to use the cleaning and plotting functions. It is *strongly* recommended that this data be stored locally in a centralized repository. Please refer to the package `README` more information on recommended work flows.

## Data sets
The functions in this package correspond to six federal fisheries data sets, each with its own respective `pull_()` function.

[NOAA Fisheries Vessel Trip Reports](https://www.fisheries.noaa.gov/inport/item/11489)

[NEFSC Observer at Sea](https://www.fisheries.noaa.gov/inport/item/24111)

[NEFSC Spring-Fall Bottom Trawl Survey](https://www.fisheries.noaa.gov/inport/item/22557)

[GARFO Permits Data](https://apps-garfo.fisheries.noaa.gov/permits/)

[GARFO Dealer Reported Landings](https://www.fisheries.noaa.gov/contact/greater-atlantic-regional-fisheries-office)

[NOAA Fisheries Marine Recreational Information Program](https://www.fisheries.noaa.gov/insight/marine-recreational-information-program)

### Vessel trip reports
`vtr <- pull_vtr(proj_path = my_path)`

```{r}
#| label: vtr
#| message: false
#| warning: false
#| echo: false

my_path <- paste0(box_path, "Confidential/VTR and Dealer/")
knitr::kable(species.shifts::pull_vtr(proj_path = my_path) |>
  head(n = 5))
```

### Fisheries Observer
`observer <- pull_observer(proj_path = my_path)`

```{r}
#| label: observer
#| message: false
#| warning: false
#| echo: false
#|
my_path <- paste0(box_path, "Confidential/Observer/DR25-200_Mills_Allyn.xlsx")
knitr::kable(species.shifts::pull_observer(proj_path = my_path) |>
  head(n = 5))
```

### NEFSC Bottom Trawl
`nefsc <- pull_nefsc(proj_path = my_path)`
```{r}
my_path <- paste0(box_path, "Non-confidential/NEFSC Trawl/survdat_lw.rds")
knitr::kable(species.shifts::pull_nefsc(proj_path = my_path) |>
  head(n = 5))
```

### GARFO Federal Permits
`permits <- pull_permits(proj_path = my_path)`
```{r}
#| label: permits
#| message: false
#| warning: false
#| echo: false

knitr::kable(readr::read_csv(
  here::here("permits_data.csv")
) |>
  head(n = 5))
```

### GARFO Dealer-reported landings
`landings <- pull_landings(proj_path = my_path)`
```{r}
#| label: landings
#| message: false
#| warning: false
#| echo: false

my_path <- paste0(box_path, "Confidential/VTR and Dealer/Dealer Data 2025.xlsx")
knitr::kable(species.shifts::pull_garfo_landings(proj_path = my_path, sheet = 1, skip = 9) |>
  head(n = 5))
```

### Marine Recreational Information Program
#### Directed trips
`mrip_directed_trips <- pull_mrip_directed_trips()`
```{r}
#| label: mrip directed trips
#| message: false
#| warning: false
#| echo: false

knitr::kable(readr::read_csv(
  here::here("mrip_example.csv")
) |>
  head(n = 5))
```

#### Catch estimates
`mrip_catch <- pull_mrip_catch()`
```{r}
#| label: mrip catch estimates
#| message: false
#| warning: false
#| echo: false

my_path <- paste0(box_path, "Non-confidential/ACCSP/Catch_Estimates.csv")

knitr::kable(species.shifts::pull_mrip_catch(proj_path = my_path) |> head(n = 5))

```
