---
title: "Accessing and cleaning federal fisheries data"
author: "Carly Lovas"
format:
  html:
    title-block-banner: "#236192"
    grid: 
      body-width: 1500px
      sidebar-width: 450px
      margin-width: 450px
    toc: true
    self-contained: true
execute: 
  echo: false
  warning: false
  message: false
  error: true
  comment: ""
  fig-align: center
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{data_cleaning_functions}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| message: false
#| warning: false
#| echo: false

library(tidyverse)
library(species.shifts)

box_path <- "/Users/clovas/Library/CloudStorage/Box-Box/MAFMC-25 Data/"
```

## About

This package contains various functions for accessing (pulling) and cleaning federal fisheries dependent and independent data. Many of these data sets are confidential and require a formal data request submission to the appropriate governing body. Information regarding metadata and access for most of these data sets can be accessed via [inPort](https://www.fisheries.noaa.gov/inport/), see below:

> InPort is the authoritative metadata repository and data inventory platform for NOAA Fisheries and the National Ocean Service. The system supports documentation of datasets and provides tools to facilitate data discovery, public access, and responsible stewardship of scientific data within these line offices.

As such, using `species.shifts` will require access to this data in order to use the cleaning and plotting functions. It is *strongly* recommended that this data be stored locally in a centralized repository. Please refer to the package `README` more information on recommended work flows.

## Data sets

The functions in this package correspond to six federal fisheries data sets, each with its own respective `pull_()` function.

[NOAA Fisheries Vessel Trip Reports](https://www.fisheries.noaa.gov/inport/item/11489)

[NEFSC Observer at Sea](https://www.fisheries.noaa.gov/inport/item/24111)

[NEFSC Spring-Fall Bottom Trawl Survey](https://www.fisheries.noaa.gov/inport/item/22557)

[GARFO Permits Data](https://apps-garfo.fisheries.noaa.gov/permits/)

[GARFO Dealer Reported Landings](https://www.fisheries.noaa.gov/contact/greater-atlantic-regional-fisheries-office)

[NOAA Fisheries Marine Recreational Information Program](https://www.fisheries.noaa.gov/insight/marine-recreational-information-program)

------------------------------------------------------------------------

### Vessel trip reports

`vtr <- pull_vtr(proj_path = my_path)`

```{r}
#| label: vtr
#| message: false
#| warning: false
#| echo: false

my_path <- paste0(box_path, "Confidential/VTR and Dealer/")
knitr::kable(species.shifts::pull_vtr(proj_path = my_path) |>
               select(!sub_trip_id) |> 
               head(n = 3))
```

\*Trip ID information removed from demo to maintain confidentiality.

### Fisheries Observer

`observer <- pull_observer(proj_path = my_path)`

```{r}
#| label: observer
#| message: false
#| warning: false
#| echo: false
#|
my_path <- paste0(box_path, "Confidential/Observer/DR25-200_Mills_Allyn.xlsx")
knitr::kable(species.shifts::pull_observer(proj_path = my_path) |>
               ungroup() |>
               select(!link3) |>
               head(n = 3))
```

\*Trip ID information removed from demo to maintain confidentiality.

### NEFSC Bottom Trawl

`nefsc <- pull_nefsc(proj_path = my_path)`

```{r}
my_path <- paste0(box_path, "Non-confidential/NEFSC Trawl/survdat_lw.rds")
knitr::kable(species.shifts::pull_nefsc(proj_path = my_path) |>
  head(n = 3))
```

### GARFO Federal Permits

`permits <- pull_permits(proj_path = my_path)`

```{r}
#| label: permits
#| message: false
#| warning: false
#| echo: false

knitr::kable(readr::read_csv(
  here::here("permits_data.csv")
) |>
  head(n = 3))
```

### GARFO Dealer-reported landings

`landings <- pull_landings(proj_path = my_path)`

```{r}
#| label: landings
#| message: false
#| warning: false
#| echo: false

my_path <- paste0(box_path, "Confidential/VTR and Dealer/Dealer Data 2025.xlsx")
knitr::kable(species.shifts::pull_garfo_landings(proj_path = my_path, sheet = 1, skip = 9) |>
               filter(is.na(confidential)) |>
               head(n = 3))
```

### Marine Recreational Information Program

#### Directed trips

`mrip_directed_trips <- pull_mrip_directed_trips()`

```{r}
#| label: mrip directed trips
#| message: false
#| warning: false
#| echo: false

knitr::kable(readr::read_csv(
  here::here("mrip_example.csv")
) |>
  head(n = 3))
```

#### Catch estimates

`mrip_catch <- pull_mrip_catch()`

```{r}
#| label: mrip catch estimates
#| message: false
#| warning: false
#| echo: false

my_path <- paste0(box_path, "Non-confidential/ACCSP/Catch_Estimates.csv")

knitr::kable(species.shifts::pull_mrip_catch(proj_path = my_path) |> head(n = 3))

```

## Plotting

Once the data has been pulled and saved to the local environment, each data set has a respective `plot_()` or `map_()` function. More about those functions [here](https://carlylovas.github.io/species.shifts/vignettes/plotting_functions.html).
