---
title: "Accessing and cleaning federal fisheries data"
author: "Carly Lovas"
format:
  html:
    title-block-banner: "#236192"
    grid: 
      body-width: 1500px
      sidebar-width: 450px
      margin-width: 450px
    toc: true
    self-contained: true
execute: 
  echo: false
  warning: false
  message: false
  error: true
  comment: ""
  fig-align: center
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{data_cleaning_functions}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| message: false
#| warning: false
#| echo: false

library(tidyverse)
library(here)
library(species.shifts)

# box_path <- "/Users/clovas/Library/CloudStorage/Box-Box/MAFMC-25 Data/"
```

## About

This package contains various functions for accessing (pulling) and cleaning federal fisheries dependent and independent data. Many of these data sets are confidential and require a formal data request submission to the appropriate governing body. Information regarding metadata and access for most of these data sets can be accessed via [inPort](https://www.fisheries.noaa.gov/inport/), see below:

> InPort is the authoritative metadata repository and data inventory platform for NOAA Fisheries and the National Ocean Service. The system supports documentation of datasets and provides tools to facilitate data discovery, public access, and responsible stewardship of scientific data within these line offices.

As such, using `species.shifts` will require access to this data in order to use the cleaning and plotting functions. It is *strongly* recommended that this data be stored locally in a centralized repository. Please refer to the package `README` more information on recommended work flows.

## Data sets

The functions in this package correspond to six federal fisheries data sets, each with its own respective `pull_()` function.

[NOAA Fisheries Vessel Trip Reports](https://www.fisheries.noaa.gov/inport/item/11489)

[NEFSC Observer at Sea](https://www.fisheries.noaa.gov/inport/item/24111)

[NEFSC Spring-Fall Bottom Trawl Survey](https://www.fisheries.noaa.gov/inport/item/22557)

[GARFO Permits Data](https://apps-garfo.fisheries.noaa.gov/permits/)

[GARFO Dealer Reported Landings](https://www.fisheries.noaa.gov/contact/greater-atlantic-regional-fisheries-office)

[NOAA Fisheries Marine Recreational Information Program](https://www.fisheries.noaa.gov/insight/marine-recreational-information-program)

------------------------------------------------------------------------

### Vessel trip reports

`vtr <- pull_vtr(proj_path = my_path)`

```{r}
#| label: vtr
#| message: false
#| warning: false
#| echo: false

# my_path <- paste0(box_path, "Confidential/VTR and Dealer/")
# knitr::kable(species.shifts::pull_vtr(proj_path = my_path) |>
#                select(!sub_trip_id) |> 
#                head(n = 3))

knitr::kable(read.csv(here::here("data","examples","vtr.csv")))
```

\*Trip ID information removed from demo to maintain confidentiality.

**How it works:** The vessel trip report data was received as a ZIP folder with several CSV files, a separte file for each year of interest. This function reads these files and combines them into a single data frame. For consistency across each file, `vtrserno` is converted into a character vector, while `year`, `sub_trip_id`, `calc_lat_deg`, `calc_lat_min`, `calc_lat_sec`, `calc_lon_deg`, `calc_lon_min`, `calc_lon_sec`, and `calc_inshr_area` are all forced as numerics.

`calc_lat_deg`, `calc_lat_min`, `calc_lat_sec` and converted and combined to create a decimal degree measurement for latitude, and the same is done for longitude. From there, any states outside GARFO jurisdiction are removed (Texas, Alabama, etc.) and trips are filtered to the area between -60 and -80 degrees west and 20 degrees north.

The data that is returned includes all that is shown above, and `sub_trip_id` as the unique trip identifier.

### Fisheries Observer

`observer <- pull_observer(proj_path = my_path)`

```{r}
#| label: observer
#| message: false
#| warning: false
#| echo: false
#|
# my_path <- paste0(box_path, "Confidential/Observer/DR25-200_Mills_Allyn.xlsx")
# knitr::kable(species.shifts::pull_observer(proj_path = my_path) |>
#                ungroup() |>
#                select(!link3) |>
#                head(n = 3))

knitr::kable(read.csv(here::here("data","examples","obs.csv")))
```

\*Trip ID information removed from demo to maintain confidentiality.

**How it works:** The fisheries observer information was received from the **Northeast Fisheries Science Center* as a Microsoft Excel Workbook. The data spans 1989 to 2024 and contains trip, haul and catch data from that time frame. The haul data is broken up into two separate sheets within the excel workbook. Catch data is stored within individual sheets for each year of the data. All trip data is contained within one sheet. 

From the `readxl` package, `excel_sheets()` to read in the entire workbook, and map `read_excel()` to read in each sheet. The two haul data sheets are combined to create a singular data frame. Catch data from 1989-1995 are missing a year column, and so one is created from the link column in order to combine these years with subsequent years of data. Haul data and catch data are joined by the `link3` column. 

Because coordinates are recorded at different points depending on the type of gear used on the trip, an intermediate data set is created to capture the proper recorded coordinates to each trip. This list is then joined back to the combined catch-haul data set to ensure accurate coordinates based on trip type. 


### NEFSC Bottom Trawl

`nefsc <- pull_nefsc(proj_path = my_path)`

```{r}
#| warning: false
# my_path <- paste0(box_path, "Non-confidential/NEFSC Trawl/survdat_lw.rds")
# knitr::kable(species.shifts::pull_nefsc(proj_path = my_path) |>
#   head(n = 3))

knitr::kable(read.csv(here::here("data","examples","nefsc.csv")))
```

**How it works:** The trawl data is received from the Northeast Fisheries Science Center as a .Rdata file pulled from an SQL database. `survdat` is extracting from the larger list of data and is the basis of the data used here. `survdat` is combined with a previously built species list to add the common names of survey species. 

From there, a unique tow ID is built based on the cruise, station, and strata surveyed. A date column is added based on the estimate month and day of the survey. 

Observations where there is a mismatch in abundance and biomass are revised so that when biomass is 0 but abundance is greater than 0, the biomass is recorded as 0.0001 kg and when abundance is 0 but biomass is greather than 0, the abundance is recorded as 1. 

Strata and species not regularly sampled are then filtered out from the data, and the data subset to begin at 1970, when the trawl began to run consistently.

### GARFO Federal Permits

`permits <- pull_permits(proj_path = my_path)`

```{r}
#| label: permits
#| message: false
#| warning: false
#| echo: false

# knitr::kable(readr::read_csv(
#   here::here("vignettes","permits_data.csv")
# ) |>
#   head(n = 3))

knitr::kable(read.csv(here::here("data","examples","permits.csv")))
```

**How it works:** 

### GARFO Dealer-reported landings

`landings <- pull_landings(proj_path = my_path)`

```{r}
#| label: landings
#| message: false
#| warning: false
#| echo: false

# my_path <- paste0(box_path, "Confidential/VTR and Dealer/Dealer Data 2025.xlsx")
# knitr::kable(species.shifts::pull_garfo_landings(proj_path = my_path, sheet = 1, skip = 9) |>
#                filter(is.na(confidential)) |>
#                head(n = 3))
knitr::kable(read.csv(here::here("data","examples","landings.csv")))

```

### Marine Recreational Information Program

#### Directed trips

`mrip_directed_trips <- pull_mrip_directed_trips()`

```{r}
#| label: mrip directed trips
#| message: false
#| warning: false
#| echo: false

# knitr::kable(readr::read_csv(
#   here::here("vignettes","mrip_example.csv")
# ) |>
#   head(n = 3))

knitr::kable(read.csv(here::here("data","examples","directedtrips.csv")))

```

#### Catch estimates

`mrip_catch <- pull_mrip_catch()`

```{r}
#| label: mrip catch estimates
#| message: false
#| warning: false
#| echo: false

# my_path <- paste0(box_path, "Non-confidential/ACCSP/Catch_Estimates.csv")
# knitr::kable(species.shifts::pull_mrip_catch(proj_path = my_path) |> head(n = 3))

knitr::kable(read.csv(here::here("data","examples","catch.csv")))


```

## Plotting

Once the data has been pulled and saved to the local environment, each data set has a respective `plot_()` or `map_()` function. More about those functions [here](https://carlylovas.github.io/species.shifts/articles/plotting_functions.html).
